The low-frequency environment of the Murchison Widefield Array: radio-frequency interference analysis and mitigation

### 2014-11-24 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Implemented some final comments of Melanie and some rewording of ADC discussions

\begin{document}
\begin{abstract}
The Murchison Widefield Array (MWA) is a new low-frequency interferometric radio telescope built in Western Australia at one of the locations of the future Square Kilometre Array (SKA). We describe the automated radio-frequency interference (RFI) detection strategy implemented for the MWA, which is based on the \textsc{aoflagger} platform, and present 72--231-MHz RFI statistics from 10 observing nights. RFI detection removes 1.1\% of the data. RFI from digital TV (DTV) is observed 3\% of the time due to occasional ionospheric or atmospheric propagation. After RFI detection and excision, almost all data can be calibrated and imaged without further RFI mitigation efforts, including observations within the FM and DTV bands. The results are compared to a previously published Low-Frequency Array (LOFAR) RFI survey. The remote location of the MWA results in a substantially cleaner RFI environment compared to LOFAR's radio environment, but adequate detection of RFI is still required before data can be analysed. We include specific recommendations designed to make the SKA more robust to RFI, including: the availability of sufficient computing power for RFI detection; analogue-digital converters with sufficient dynamic range; a smooth band-pass response; and the capability of RFI detection at high time and frequency resolution (second and kHz-scale respectively).
\end{abstract}

@@

After RFI detection and excision, power spectra from both MWA and LOFAR show smooth curves without significant leakage, and the data are clean enough for further processing and imaging. From the difference between 1.65\% or 3.18\% loss of data at the MWA and LOFAR sites respectively, it is clear that the impact of RFI is smaller for the MWA, but RFI mitigation is still required with similar per-sample computational requirements as for LOFAR. When comparing the 115--163~MHz range, the RFI occupancy levels are small and the RFI flagging is effective for both sites. Therefore, the MWA's larger radio-quiet zone does not provide a significant benefit at these frequencies. However, a benefit of the remote radio-quiet site, is that the MWA can observe in the 88--108~MHz FM-station and 174-230 MHz DTV bands. Initial MWA experiments have confirmed the availability of these bands for science \citep{mckinley-moon-2013,hurley-walker-mwacs-2014}. Moreover, for both LOFAR and MWA, a primary science driver is the detection of signals from the EoR. The required integration time for such a detection of approximately a hundred nights has currently not been achieved, and it could be that RFI becomes problematic when reaching lower noise levels. Initial results with LOFAR predict that RFI will not prevent such a detection \citep{ncp-eor-yatawatta, offringa-rfi-distributions}.

@@

SKA-low will be built at the same location as the MWA, and hence several lessons can be learned. First of all, it is clear that the SKA will need to be able to handle some amount of RFI. This requires computational power to perform the RFI detection, analogue-digital converters (ADCs) with sufficient dynamic range, and a signal path with headroom sufficient to avoid saturation by RFI. Second, \textsc{cotter} relies on the fact that a single computing node can still hold a reasonable amount of MWA data in memory for the RFI detection, but because of SKA's large number of elements and high time and frequency resolution, this will be a more challenging problem. Thirdly, flagging MWA data is slightly complicated due to the first poly-phase filter and digital gains. The extra per-subband gain correction that is required for the MWA makes the detection less stable, and the sub-band bandpass makes it harder to recognize RFI patterns in frequency direction. Therefore, for accurate RFI detection it is best to have a smooth response over a large instantaneous bandwidth. Finally, the presence of short and spectrally-narrow RFI events confirms that detection at high time and frequency resolution improves accuracy.

>>>>

\begin{document}
\begin{abstract}
The Murchison Widefield Array (MWA) is a new low-frequency interferometric radio telescope built in Western Australia at one of the locations of the future Square Kilometre Array (SKA). We describe the automated radio-frequency interference (RFI) detection strategy implemented for the MWA, which is based on the \textsc{aoflagger} platform, and present 72--231-MHz RFI statistics from 10 observing nights. RFI detection removes 1.1\% of the data. RFI from digital TV (DTV) is observed 3\% of the time due to occasional ionospheric or atmospheric propagation. After RFI detection and excision, almost all data can be calibrated and imaged without further RFI mitigation efforts, including observations within the FM and DTV bands. The results are compared to a previously published Low-Frequency Array (LOFAR) RFI survey. The remote location of the MWA results in a substantially cleaner RFI environment compared to LOFAR's radio environment, but adequate detection of RFI is still required before data can be analysed. We include specific recommendations designed to make the SKA more robust to RFI, including: the availability of sufficient computing power for RFI detection; accounting for RFI in the receiver design; a smooth band-pass response; and the capability of RFI detection at high time and frequency resolution (second and kHz-scale respectively).
\end{abstract}

@@

After RFI detection and excision, power spectra from both MWA and LOFAR show smooth curves without significant leakage, and the data are clean enough for further processing and imaging. From the difference between 1.65\% or 3.18\% loss of data at the MWA and LOFAR sites respectively, it is clear that the impact of RFI is smaller for the MWA, but RFI mitigation is still required with similar per-sample computational requirements as for LOFAR. In both cases, RFI occupancy levels are small and RFI flagging is effective. Nevertheless, a benefit of MWA's remote radio-quiet site is that it allows observations in the 88--108~MHz FM-station and 174-230 MHz DTV bands. Initial MWA experiments have confirmed the availability of these bands for science \citep{mckinley-moon-2013,hurley-walker-mwacs-2014}. Moreover, for both LOFAR and MWA, a primary science driver is the detection of signals from the EoR. The required integration time for such a detection of approximately a hundred nights has currently not been achieved, and it could be that RFI becomes problematic when reaching lower noise levels. Initial results with LOFAR predict that RFI will not prevent such a detection \citep{ncp-eor-yatawatta, offringa-rfi-distributions}.

@@

SKA-low will be built at the same location as the MWA, and hence several lessons can be learned. First of all, it is clear that the SKA will need to be able to handle some amount of RFI. This requires computational power to perform the RFI detection, and a receiver signal path with headroom sufficient to avoid saturation by RFI. Second, \textsc{cotter} relies on the fact that a single computing node can still hold a reasonable amount of MWA data in memory for the RFI detection, but because of SKA's large number of elements and high time and frequency resolution, this will be a more challenging problem. Thirdly, flagging MWA data is slightly complicated due to the first poly-phase filter and digital gains. The extra per-subband gain correction that is required for the MWA makes the detection less stable, and the sub-band bandpass makes it harder to recognize RFI patterns in frequency direction. Therefore, for accurate RFI detection it is best to have a smooth response over a large instantaneous bandwidth. Finally, the presence of short and spectrally-narrow RFI events confirms that detection at high time and frequency resolution improves accuracy.
